---
title: "ERHS 535 Final Project: Prediction of Human Cases of West Niles Virus in Southern California by Precipitation and Temperature (2006-2010)"
author: "Devin, Anastasia, Grant, Ana"
date: "11/7/2018"
output: pdf_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# 1. Packages used in the final project

```{r, Packages needed}
library(dplyr)
library(stringr)
library(tidyselect)
library(ggplot2)
library(rmarkdown)
library(RColorBrewer)
library(knitr)
library(forcats)
library(readr)
library(tidyr)
library(broom)
library(purrr)
library(scales)
library(lubridate)
library(viridis)
library(RColorBrewer)
library(scales)
library(tidyverse)
library(tigris)
library(sf)
library(cowplot)
library(gridExtra)
```

# 2. Datasets description

Different data sets were collected from CHHS and CDC Wonder. The report of human cases for California, had data arrange by week and county for the years 2006 to 2018. The data is available in:

[CHHS] <https://data.chhs.ca.gov/dataset/west-nile-virus-cases-2006-present/resource/6ef33c1b-9f54-49f2-a92e-51a1b78f0a06>

Daily precipitation and temperature in California were downloaded from CDC Wonder website. Every year was pulled independently. The request form might be found in the following website.

[CDC Wonder]<https://wonder.cdc.gov/>

```{r, uploading data sets used, human cases and weather data}
report <-  read_csv( "../Data/wnv_human_cases.csv")
temp_2006 <- read_csv("../Data/temp/temp_2006.csv")
temp_2007 <- read_csv("../Data/temp/temp_2007.csv")
temp_2008 <- read_csv("../Data/temp/temp_2008.csv")
temp_2009 <- read_csv("../Data/temp/temp_2009.csv")
temp_2010 <- read_csv("../Data/temp/temp_2010.csv")
precip_2006 <- read_csv("../Data/precip/precip_2006.csv")
precip_2007 <- read_csv("../Data/precip/precip_2007.csv")
precip_2008 <- read_csv("../Data/precip/precip_2008.csv")
precip_2009 <- read_csv("../Data/precip/precip_2009.csv")
precip_2010 <- read_csv("../Data/precip/precip_2010.csv")

```

# 3. Tyding and organizing the original data sets

## 3.1. Human cases dataset

The main objective was change the *weekly* report to a *year* report for the number of cases by county

```{r, cleaning human cases dataset}
cases <- report %>% 
  select(Year, 
         county = County, 
         positive_cases = "Positive Cases") %>% 
  mutate(date = Year) %>% 
  filter(date %in% c("2006", "2007", "2008", "2009", "2010")) %>% 
  group_by(date, county) %>% 
  summarize(positive_cases = sum(positive_cases)) %>% 
  ungroup %>% 
  arrange(date)

kable(head(cases))
kable(tail(cases))

```

## 3.2. Temperature dataset

The main objective was join all the temperature data in one data set and select the relevant variables.

```{r, joining all temperature files}

temp <- rbind(temp_2006, temp_2007, temp_2008, temp_2009, temp_2010)

temp <- temp %>% 
  select(County, "County Code", "Month Day, Year Code", "Day of Year", 
         "Avg Daily Max Air Temperature (F)", 
         "Avg Daily Min Air Temperature (F)") %>% 
  rename(county = County,
         fips = "County Code",
         date = "Month Day, Year Code",
         day_year = "Day of Year",
         max_temp = "Avg Daily Max Air Temperature (F)",
         min_temp_f = "Avg Daily Min Air Temperature (F)") %>% 
  mutate(date = mdy(date)) 
kable(head(temp))

```

## 3.3. Precipitation dataset

Same as the previous objecyive, precipitation datasets were joined and the relevant information was selected.

```{r, joining all precipitation files}

precip <- rbind(precip_2006, precip_2007, precip_2008, precip_2009, precip_2010)

precip <- precip %>%
  select(County, "Month Day, Year Code", "Avg Daily Precipitation (mm)")%>%
  rename(county = County,
         date = "Month Day, Year Code",
         avg_precip = "Avg Daily Precipitation (mm)") %>% 
  mutate(date = mdy(date))

kable(head(precip))

```

# 4. Merging cleaned datasets

Two approaches were considered. 

- First, precipitation and temperature datasets were merged. Right after that averages by year were calculated.

- Second,the yearly human West Niles Virus cases dataset was joined with weather data.

## 4.1. Weather data

In this step, temperature and precipitation data sets were merged.Addionally, calculation of yearly averages

```{r, merging precipitations and temperature and obtaining average precipitation and temperature by year}

ca_weather <- merge(temp, precip, by = c("county", "date")) %>% 
  select(county, date, fips, avg_precip, max_temp) %>% 
  separate(county, c("county", "state"), sep = " County, CA") %>% 
  select(county, date, fips, avg_precip, max_temp) %>% 
  mutate(date = year(date)) %>% 
  group_by(county, fips, date) %>% 
  summarise(avg_precip = mean(avg_precip), avg_max_temp = mean(max_temp)) %>% 
  ungroup 

kable(tail(ca_weather))

```

## 4.2. Cases and weather data

In this step cases and weather dataset were joined

```{r, joining cases of west niles and california weather}

ca_weather_cases <- full_join(ca_weather, cases, by = c('date', 'county')) %>%
  mutate(positive_cases = ifelse(!is.na(positive_cases), positive_cases, 0)) %>% 
  arrange(positive_cases) 

kable(head(ca_weather_cases))

```

# 5. Creating the final data set

To create the final dataset and in order to be able to create a Shinny App, geographical data from California counties was joined with human cases and weather data (joined in part 4.2).In addition, for the structure of the data just counties that are part of Southern California will be analyzed.

```{r, final data set, joining geographical data with cases and weather in California, message = FALSE}

ca_counties <- counties(state = "CA", cb = TRUE, class = "sf")

ca_county_cases <- ca_counties %>% 
  mutate(fips = paste(STATEFP, COUNTYFP, sep = "")) %>% 
  full_join(ca_weather_cases, by = "fips") %>% 
  filter(county %in% c("Imperial" , "Kern", "Los Angeles", "Orange",
                       "Riverside", "San Bernardino", "San Diego", 
                       "San Luis Obispo", "Santa Barbara", "Ventura"))

#kable(head(ca_county_cases, 3)) 

unique(ca_county_cases$county)
unique(ca_county_cases$date)
```

